Wait-free Queues with Polygarithmic Step Complexity

### 2022-05-20 article/psuedocode.tex
polish and add proof for some lemmas

\begin{definition}[Block]
A block is an object that stores some statistics described in Algorithm Queue. It implicitly shows a set of operations. The set of operations of block $b$ are the operations in the leaf subblocks of $b$. We show the set of operations of block $b$, set of blocks $B$ by $ops(b)$, $ops(B)$. We also say $b$ contains $op$ if $op\in ops(b)$.
\end{definition}

@@

\begin{lemma}
  Each block is not duplicated.
\end{lemma}

@@

\begin{lemma}[createBlock] \label{lem::createBlock}
  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $b$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
\end{lemma}
\begin{proof}
 WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the established operations at time $t$ by definition of subblock. Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}

@@

\begin{lemma}[Double Refresh] \label{doubleRefresh}
  Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process. Let $t_1$ be the time $r_1$ is invoked and $t_2$ be the time $r_2$ terminated. After $r_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
\end{lemma}
%\begin{proof}
%  
%
%If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}. If not, then there is another successful instance of \nf{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}. $R_1^\prime$ may or may not have propagated some subblocks of \nf{new} to \nf{n}. It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}). If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds. If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}. Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$. So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$. This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds. See Figure~\ref{}.
%\end{proof}

@@

\begin{corollary}[Propagate Step] \label{doublyRefresh}
All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s established blocks after line~\ref{secondRefresh}.
\end{corollary}
%\begin{proof}
%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35. By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}. \nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well. First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursively. If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails. It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
%\end{proof}

@@

\pagebreak
\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.

@@

\begin{theorem}
The queue implementation is linearizable.
\end{theorem}
%\begin{proof}
%  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
%  \begin{enumerate}
%    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}. The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
%    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$. \\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered. It remains to prove  Dequeue $d$ returns the correct response according to the linearization order. By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly. If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
%  \end{enumerate} 
%\end{proof}

>>>>

\framebox[1.1\width]{TEST} How to merge notions of blocks and operations? block b $\in$ block c means b is subblock of c. block b $\in$ set B means b is in B. Merge these two to have shorter formulaes.

\begin{definition}[Block]
A block is an object storing some statistics described in Algorithm Queue. It implicitly shows a set of operations. The set of operations of block $b$ are the operations in the leaf subblocks of $b$. We show the set of operations of block $b$, set of blocks $B$ by $ops(b)$, $ops(B)$. We also say $b$ contains $op$ if $op\in ops(b)$.
\end{definition}

@@

\begin{lemma}
  Each block is not duplicated.
\end{lemma}
\begin{proof}
  \nf{n.CreateBlock(i)} is successful for \nf{i=1,2,3,...} exactly one time and by definition of \nf{CreateBlock} the range of their blocks are disjoint so in propagation procedure on block can not be propagated twice (reference to propagation).
\end{proof}

@@

\begin{lemma}[createBlock] \label{lem::createBlock}
  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
\end{lemma}
\begin{proof}
From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}. It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$. If $b$ is $\in$ \nf{EST\sub{n.left,t$^\prime$}} it is simple to see. Else we can see from the code of the \nf{CreateBlock} that \nf{new} contains it. Right child is the same.
% WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.  Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$. By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}. Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10. 
By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}

@@

\begin{lemma}[Double Refresh] \label{doubleRefresh}
  Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process. Let $t_1$ be the time $r_1$ is invoked and $t_2$ be the time $r_2$ terminated. After $r_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
\end{lemma}
\begin{proof}

If Line 313 of $r_1$ or $r_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}. If not, then there is another successful instance of \nf{n.Refresh()} $r_1^\prime$ by Lemma~\ref{lem:falseRefresh} for $r_2$. From $t_{i_{r_2}^\prime} > t_1$ and Lemma~\ref{lem::trueRefresh}, \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
%It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}). If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds. If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}. Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$. So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$. This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds. See Figure~\ref{}.
\end{proof}

@@

\begin{corollary}[Propagate Step] \label{doublyRefresh}
All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s established blocks after line~\ref{secondRefresh}.
\end{corollary}
\begin{proof}
Lines \ref{firstRefresh} and \ref{secondRefresh} satisfy the preconditions of Lemma \ref{doubleRefresh}.
\end{proof}

@@

\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.

@@

\begin{theorem}[Main]
The queue implementation is linearizable.
\end{theorem}
%\begin{proof}
%  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
%  \begin{enumerate}
%    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}. The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
%    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$. \\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered. It remains to prove  Dequeue $d$ returns the correct response according to the linearization order. By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly. If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
%  \end{enumerate} 
%\end{proof}

\begin{lemma}[Time analysis]
  \nf{n.GetEnq(b,i), n.Index(b,i)} take $O(\log^2 p)$ steps. Search in the root may take $O(\log Q+ p^2)$ steps. Helping is done every $p^2$ block appended to the root and takes $p\times \log^2p$ steps. Amortized time consumed for helping by each process is $O(\log^2 p)$.
\end{lemma}
